### 2019-11-27 2-Chapters/4-Chapter/CrownCom_17/IoT_slotted.tex
Removed raw inclusion of https://hal.inria.fr/hal-02006825 https://hal.inria.fr/hal-02049824v2 https://hal.archives-ouvertes.fr/hal-01575419v2
I fixed a lot of small typing mistakes, reported by Nathalie Mitton.
- Thanks to her!
- Shame on me haha

\graphicspath{{2-Chapters/4-Chapter/CrownCom_17.git/}}

@@

We note that MAB learning has also started to be applied to also optimize the PHY layer parameters, see for instance \cite{KerkoucheAlami18} which followed our article \cite{Bonnefoi17},
but in this thesis we chose to restrict to the spectrum access problem and thus we only MAB models where the arms correspond to orthogonal frequency bands.

@@

This problem is a special case of the stochastic MAB with
% , where the sequence of rewards drawn from a given arm $k$ is assumed to be  \emph{i.i.d.}, under some distribution $\nu_k$, that has a mean $\mu_k$.
% Several types of reward distributions have been considered in the literature, and rewards are binary in our model.
% We consider only
Bernoulli distributions.
% , in which $r_k(t) \sim \mathrm{Bern}(\mu_k)$, that is, $r_k(t) \in \{0,1\}$ and $\mathbb{P}(r_k(t) = 1) = \mu_k \in [0,1]$.
Contrarily to many previous work done in the CR field and for Opportunistic Spectrum Access \cite{Jouini10,Jouini12},
the reward $r(t)$ does \emph{not} come from a sensing phase before sending the $t$-th message, as it would do for any ``listen-before-talk'' model.
In our model, rewards rather come from receiving or not an acknowledgement from the gateway, between the $t$-th and $t+1$-th messages. A reward of one indicates that the acknowledgement was received on time, and a reward of zero indicates the opposite.

@@

All dynamic devices follow the same policy in this case, so the probability of transmitting at that time in channel $k$ for any dynamic device is $p / K$, and there are $D-1$ other dynamic devices.
As they are independent, the probability that no other dynamic device sent in $i$
% $\Pr(\text{no other dynamic device sent in}\;i)$,
is $q = \Pr(\bigcap_{k=1}^{D-1} \text{device } k \text{ did not send in } i) = \prod_{k=1}^{D-1} \Pr(\text{device } k \text{ did not send in } i)$. And $\Pr(\text{device } k \text{ sent in } i) = p \times 1 / K$, by uniform choice on channels and the Bernoulli emission hypothesis. So $q = \prod_{k=1}^{D-1} (1 - p/K) = (1-p/K)^{D-1}$. Thus we can conclude,
%
\begin{align}\label{eq:41:strategynaive}
    \Pr(\text{success}|\text{sent})
    & = \sum_{k=1}^{K} \underbrace{(1 - p / K)^{D-1}}_{\text{No other dynamic device}} \times \underbrace{(1-p)^{S_k}}_{\text{No static device}} \times\; \frac{1}{K} \nonumber \\
    & = \frac{1}{K} \left(1-\frac{p}{K}\right)^{D-1} \sum_{k=1}^{K} (1-p)^{S_k} .
\end{align}
%
This probability \eqref{eq:41:strategynaive} is constant wrt time, and easy to compute numerically.
Comparing the successful transmission rate of any policy against this naive policy is important, as any efficient learning algorithm should outperform it,
maybe after a long enough initial learning period.

@@

While the stochastic MAB model has been used to describe some aspects of Cognitive Radio systems, it is in principle not suitable for our IoT model, due to the non-stationarity of the channels occupancy, caused partly by the learning policy used by dynamic devices, but mainly by their random activation processes.
%
In our model, every dynamic device implements its own learning algorithm, \emph{independently}.
For one device, the time $t$ refers to the number of time it accessed the network (following its Bernoulli transmission process, \ie, its duty cycle), \emph{not} the total number of time slots from the beginning, as rewards are only obtained after a transmission, and IoT devices only transmit sporadically, due to low transmission duty cycles.

@@

\paragraph{First results.}
%
Figure~\ref{fig:41:from10to100} presents the successful transmission rate as a function of time.
The two MAB algorithms, \UCB{} and Thompson Sampling (TS), are compared against the naive random policy (which are outperformed easily by the MAB algorithms), and the two (optimal and greedy) oracle policies (which outperform slightly the MAB algorithms).
The results are displayed when $10\%$, $30\%$, $50\%$ and $100\%$ of the traffic is generated by dynamic devices.

>>>>

\graphicspath{{2-Chapters/4-Chapter/CrownCom_17/}}

@@

We note that MAB learning has also started to be applied to also optimize the PHY layer parameters, see for instance \cite{KerkoucheAlami18} which followed our article \cite{Bonnefoi17},
but in this thesis we chose to restrict to the spectrum access problem and thus we only consider MAB models where the arms correspond to orthogonal frequency bands.

@@

This problem is a special case of the stochastic MAB with
% , where the sequence of rewards drawn from a given arm $k$ is assumed to be  \emph{i.i.d.}, under some distribution $\nu_k$, that has a mean $\mu_k$.
% Several types of reward distributions have been considered in the literature, and rewards are binary in our model.
% We consider only
Bernoulli distributions.
% , in which $r_k(t) \sim \mathrm{Bern}(\mu_k)$, that is, $r_k(t) \in \{0,1\}$ and $\mathbb{P}(r_k(t) = 1) = \mu_k \in [0,1]$.
Contrarily to many previous works done in the CR field and for Opportunistic Spectrum Access \cite{Jouini10,Jouini12},
the reward $r(t)$ does \emph{not} come from a sensing phase before sending the $t$-th message, as it would do for any ``listen-before-talk'' model.
In our model, rewards rather come from receiving or not an acknowledgement from the gateway, between the $t$-th and $t+1$-th messages. A reward of one indicates that the acknowledgement was received on time, and a reward of zero indicates the opposite.

@@

All dynamic devices follow the same policy in this case, so the probability of transmitting at that time in channel $k$ for any dynamic device is $p / K$, and there are $D-1$ other dynamic devices.
As they are independent, the probability that no other dynamic device sent in $k$
% $\Pr(\text{no other dynamic device sent in}\;i)$,
is $q = \Pr(\bigcap_{j=1}^{D-1} \text{device } j \text{ did not send in } k) = \prod_{j=1}^{D-1} \Pr(\text{device } j \text{ did not send in } k)$. And $\Pr(\text{device } j \text{ sent in } k) = p \times 1 / K$, by uniform choice on channels and the Bernoulli emission hypothesis. So $q = \prod_{j=1}^{D-1} (1 - p/K) = (1-p/K)^{D-1}$. Thus we can conclude,
%
\begin{align}\label{eq:41:strategynaive}
    \Pr(\text{success}|\text{sent})
    & = \sum_{k=1}^{K} \underbrace{(1 - p / K)^{D-1}}_{\text{No other dynamic device}} \times \underbrace{(1-p)^{S_k}}_{\text{No static device}} \times\; \frac{1}{K} \nonumber \\
    & = \frac{1}{K} \left(1-\frac{p}{K}\right)^{D-1} \sum_{k=1}^{K} (1-p)^{S_k} .
\end{align}
%
This probability \eqref{eq:41:strategynaive} is constant wrt time, and easy to compute numerically.
Comparing the successful transmission rate of any policy against this naive policy is important, as any efficient learning algorithm should outperform it,
maybe after a long enough initial learning period.

@@

While the stochastic MAB model has been used to describe some aspects of Cognitive Radio systems, it is in principle not suitable for our IoT model, due to the non-stationarity of the channels occupancy, caused partly by the learning policy used by dynamic devices, but mainly by their random activation processes.
%
In our model, every dynamic device implements its own learning algorithm, \emph{independently}.
For one device, the time $t$ refers to the number of times it accessed the network (following its Bernoulli transmission process, \ie, its duty cycle), \emph{not} the total number of time slots from the beginning, as rewards are only obtained after a transmission, and IoT devices only transmit sporadically, due to low transmission duty cycles.

@@

\paragraph{First results.}
%
Figure~\ref{fig:41:from10to100} presents the successful transmission rate as a function of time.
The two MAB algorithms, \UCB{} and Thompson Sampling (TS), are compared against the naive random policy (which is outperformed easily by the MAB algorithms), and the two (optimal and greedy) oracle policies (which outperform slightly the MAB algorithms).
The results are displayed when $10\%$, $30\%$, $50\%$ and $100\%$ of the traffic is generated by dynamic devices.

### 2019-11-27 2-Chapters/4-Chapter/IEEE_WCNC__2019__Paper__BMBM/IEEE_WCNC__2019__Paper__BMBM.tex
Removed raw inclusion of https://hal.inria.fr/hal-02006825 https://hal.inria.fr/hal-02049824v2 https://hal.archives-ouvertes.fr/hal-01575419v2
I fixed a lot of small typing mistakes, reported by Nathalie Mitton.
- Thanks to her!
- Shame on me haha

\graphicspath{{2-Chapters/4-Chapter/IEEE_WCNC__2019__Paper__BMBM.git/}}

@@

Like in the previous model presented in Section~\ref{sec:4:firstModel},
regardless the type of devices, each of them follows a slotted ALOHA protocol \cite{Roberts75}, and has a probability $p>0$ to transmit a packet in a time slot.
We make the hypothesis that the transmission is successful if the channel is available, otherwise it fails upon radio collision.
The novelty compare to the previous model is that
in case of RF (up-link or down-link) collision that prevent the device to receive the \emph{Ack} from the gateway,
these devices will attempt to retransmit their packet up-to $\mathrm{MaxBackOff}$ times\footnote{~We denote it $\mathrm{MaxBackOff}$ instead of $M$ like we did in our paper \cite{Bonnefoi2019WCNC}, as $M$ is used in next Chapter~\ref{chapter:5}.},
with $\mathrm{MaxBackOff} \in\mathbb{N}^*$.
It is important to note that, every retransmission is carried out after a random back-off time, uniformly distributed in $\{0, \dots, m-1\}$, where $m \in\mathbb{N}^*$ is the length of the back-off interval
(note the difference between $m$ and $\mathrm{MaxBackOff}$).

@@

It is worth to highlight that, if we write \eqref{eq:43:decomppc1} as $p_{c1} = p_c + p_{ca} \left(1-p_c\right)$,
then it is obvious that $p_{c1}$ is always larger than $p_c$ (as $p_{ca} \left(1-p_c\right) > 0$).
But for large values of $p_c$, $p_{ca}\left(1-p_c\right) \simeq 0$ so the gap gets small,
and for small values of $p_c$ the gap is significant.
Moreover, we can verify (\eg, numerically or by differentiating)
that the gap decreases when $p_c$ increases (for fixed $N$ and $m$).
This backups mathematically the observation we made from Figure~\ref{fig:43:Approximation_m10}:
the smaller the $p_c$, the larger is the gap between $p_c$ and $p_{c1}$.
%
We interpret this fact in two different situations:
\begin{itemize}
	\item
	On the one hand, in a congested network, when devices suffer from a large probability of collision on their first transmission (\ie, $p_c$ is not so small), then $p_{c1}\simeq p_c$ and so devices cannot really hope to reduce their collision probabilities even if the use a different channel for retransmission.
	\item
	On the other hand, if $p_c$ is small enough, \ie, in a network not yet too congested, then the derivation above shows that $p_{c1} > p_c$, meaning that the possible gain of retransmitting in a different channel that the one used for the first transmission can be large, in terms of collision probability (\eg, up-to $10\%$ in this experimental setting).
	In other words, when learning can be useful (small $p_c$), learning to retransmit in a different channel can have a large impact on the global collision rate,
	thus justifying our approach.
\end{itemize}

>>>>

\graphicspath{{2-Chapters/4-Chapter/IEEE_WCNC__2019__Paper__BMBM/}}

@@

Like in the previous model presented in Section~\ref{sec:4:firstModel},
regardless the type of devices, each of them follows a slotted ALOHA protocol \cite{Roberts75}, and has a probability $p>0$ to transmit a packet in a time slot.
We make the hypothesis that the transmission is successful if the channel is available, otherwise it fails upon radio collision.
The novelty compared to the previous model is that
in case of RF (up-link or down-link) collision that prevents the devices from receiving the \emph{Ack} from the gateway,
these devices will attempt to retransmit their packet up-to $\mathrm{MaxBackOff}$ times\footnote{~We denote it $\mathrm{MaxBackOff}$ instead of $M$ like we did in our paper \cite{Bonnefoi2019WCNC}, as $M$ is used in next Chapter~\ref{chapter:5}.},
with $\mathrm{MaxBackOff} \in\mathbb{N}^*$.
It is important to note that, every retransmission is carried out after a random back-off time, uniformly distributed in $\{0, \dots, m-1\}$, where $m \in\mathbb{N}^*$ is the length of the back-off interval
(note the difference between $m$ and $\mathrm{MaxBackOff}$).

@@

It is worth highlighting that, if we write \eqref{eq:43:decomppc1} as $p_{c1} = p_c + p_{ca} \left(1-p_c\right)$,
then it is obvious that $p_{c1}$ is always larger than $p_c$ (as $p_{ca} \left(1-p_c\right) > 0$).
But for large values of $p_c$, $p_{ca}\left(1-p_c\right) \simeq 0$ so the gap gets small,
and for small values of $p_c$ the gap is significant.
Moreover, we can verify (\eg, numerically or by differentiating)
that the gap decreases when $p_c$ increases (for fixed $N$ and $m$).
This backups mathematically the observation we made from Figure~\ref{fig:43:Approximation_m10}:
the smaller the $p_c$, the larger is the gap between $p_c$ and $p_{c1}$.
%
We interpret this fact in two different situations:
\begin{itemize}
	\item
	On the one hand, in a congested network, when devices suffer from a large probability of collision on their first transmission (\ie, $p_c$ is not so small), then $p_{c1}\simeq p_c$ and so devices cannot really hope to reduce their collision probabilities even if the use a different channel for retransmission.
	\item
	On the other hand, if $p_c$ is small enough, \ie, in a network not yet too congested, then the derivation above shows that $p_{c1} > p_c$, meaning that the possible gain of retransmitting in a different channel that the one used for the first transmission can be large, in terms of collision probability (\eg, up-to $10\%$ in this experimental setting).
	In other words, when learning can be useful (small $p_c$), learning to retransmit in a different channel can have a large impact on the global collision rate,
	thus justifying our approach.
\end{itemize}

### 2019-11-27 2-Chapters/5-Chapter/ALT_2018__MPBandits/mpBandits.tex
I fixed a lot of small typing mistakes, reported by Nathalie Mitton.
- Thanks to her!
- Shame on me haha

\begin{itemize}
  \item[\modelun]
    \textbf{Simultaneous sensing and collision}: player $j$ observes  $Y_{A^j(t),t}$ \emph{and} $C^j(t)$.
    We note that this first model was never previously studied, but we do not focus on it because of its unrealistic aspect, and its simplicity (it is easier than the following model, for which we obtain good theoretical results).
  \item[\modeldeux]
    \textbf{Sensing, then collision}: player $j$ observes $Y_{A^j(t),t}$, \emph{then} observes the reward, and thus also $C^j(t)$ only if $Y_{A^j(t),t} = 1$.
    This common setup, studied for example by \cite{Anandkumar11,Rosenski16}, is relevant to model the OSA problem: the device first checks for the presence of primary users in the chosen channel,
    if this channel is free ($Y_{A^j(t),t}=1$), the transmission is successful ($r^j(t)=1$) if no collision occurs with other smart devices ($\overline{C^j(t)}$).
  \item[\modeltrois]
    \textbf{No sensing}: player $j$ only observes the reward $r^j(t)$.
    For IoT networks, this reward can be interpreted as an acknowledgement from a Base Station,
    received when a communication was successful.
    A lack of acknowledgment may be due to a collision
    with a device from the background traffic $(Y_{A^j(t),t}=0)$,
    or to a collision with one of the others players ($C^j(t)$).
    However, the sensing and collision information are censored.
    Recently, our work presented in Chapter~\ref{chapter:4} \cite{Bonnefoi17} presented the first (bandit-based) algorithmic solutions under this (harder) feedback model, in a slightly different setup, more suited to large scale IoT applications. As explained above, this third model was found to be too general to analyze mathematically, and we did not present theoretical results of convergence of the algorithms considered in Chapter~\ref{chapter:4}.
    %\footnote{\cite{Bonnefoi17} considers a different setting, more suited for large-scale IoT networks, and a future work will be to extend our work to this setting with more than $K$ players but small probability of activations.}.
\end{itemize}

@@

The regret decomposition in Lemma~\ref{lem:5:DecompositionRegret} is valid for both centralized and decentralized algorithms.
For centralized algorithms, due to the absence of collisions, \ref{eq:5:term3} is obviously zero, and \ref{eq:5:term2} is non-negative, as $N_k(T) \leq T$. For decentralized algorithms, \ref{eq:5:term3} may be significantly large, and term \ref{eq:5:term2} may be negative, as many collisions on arm $k$ may lead to $N_k(T) > T$ (which is counter intuitive with such notations).
However, a careful manipulation of this decomposition shows that the regret is always lower bounded by term \ref{eq:5:term1}.
% The Figure~\ref{fig:5:MP__M9_K9_T10000_N1000__9_algos__main_RegretCentralized____env6} in Appendix illustrate two cases of $M<K$ and $M=K$ and the different impact of the three terms \ref{eq:5:term1}, \ref{eq:5:term2} and \ref{eq:5:term3} on the regret.
The Figure~6 in Appendix~F of \cite{Besson2018ALT} illustrates two cases of $M<K$ and $M=K$, and the different impact of the three terms \ref{eq:5:term1}, \ref{eq:5:term2} and \ref{eq:5:term3} on the regret.

@@

Note that $\widetilde{\mu_k}^j(t)$ is no longer meant to be an unbiased estimate of $\mu_k$ as it also takes into account the collision information, that is present in the reward. Based on this empirical reward, one can similarly defined modified indices as
%
\begin{equation}\label{eq:5:indexTilde}
  \widetilde{U_k}^j(t) \eqdef \begin{cases}
      \widetilde{\mu_k}^j(t)  + \sqrt{  f(t) / (2N_k^j(t))}
      &\text{for } \UCB, \\
      \sup\left\{ q \in [0, 1]: N_k^j(t)\times \kl(\widetilde{\mu_k}^j(t), q) \leq f(t) \right\}
      &\text{for } \klUCB.
  \end{cases}
\end{equation}

@@

% \textbf{Index policies.}
% %
The approaches we now describe for multi-players bandits can be used in combination with any index policy (see Algorithm~\ref{algo:2:indexPolicy}), but we restrict our presentation to \UCB{} algorithms, for which strong theoretical guarantees can be obtained. In particular, we focus on two types of indices:
$\UCB_1$ \citep{Auer02}
and \klUCB{} \citep{KLUCBJournal}.
For player $j\in[M]$
%
denote $S_k^j(t) \eqdef \sum_{s=1}^t Y_{k,s} \indic(A^j(t) = k)$ the current sum of rewards obtained for arm $k$,
then $\widehat{\mu}_k^j(t) \eqdef S_k^j(t)/N_k^j(t)$ (if $N_k^j(t)\neq 0$) is the empirical mean of arm $k$, and thus one can define the index
\begin{equation}\label{eq:5:indexFor_UCB_klUCB}
  U_k^j(t) \eqdef \begin{cases}
      \widehat{\mu}_k^j(t)  + \sqrt{  f(t) / (2N_k^j(t))}
      &\text{for } \UCB, \\
      \sup\left\{ q \in [0, 1]: N_k^j(t)\times \kl(\widehat{\mu}_k^j(t), q) \leq f(t) \right\}
      &\text{for } \klUCB,
  \end{cases}
\end{equation}
where $f(t)$ is some \emph{exploration function}, usually taken to be $\log(t)$ in practice, and slightly larger in theory (\eg, $f(t)=\log(t)+3\log(\log(t))$), which ensures that  $\Pr(U_k^j(t) \geq \mu_k) \gtrsim 1 - 1/t$ (see \cite{KLUCBJournal}).
A classical (single-player) UCB algorithm aims at the arm with largest index, as presented above in Section~\ref{sub:2:IndexPolicies}.
However, if each of the $M$ players selects the arm with largest UCB, all the players will end up colliding most of the time on the best arm.
To circumvent this problem, several coordination mechanisms have emerged, that rely on \emph{ordering} the indices and targeting \emph{one of} the $M$-best indices.

@@

\paragraph{The \RandTopM{} algorithm.}
%
We precisely state the first proposal below in Algorithm~\ref{algo:5:RandTopM}.
%
\RandTopM{} is essentially a refinement over \RhoRand, to not use the indirection of ranks, and a simpler version of \MCTopM.
The difference with \MCTopM{} is that the later introduces a concept of a ``Chair'', by considering a binary ``being fixed'' state $s^j(t)$, as presented in Algorithm~\ref{algo:5:MCTopM} below.
%
In \RandTopM{}, player $j$ is always considered ``not fixed'',
and a \emph{collision always forces a uniform sampling of the next arm} from $\TopM(t)$.

@@

% for N in 4 100; do for M in 3 6 9; do DEBUG=True SAVEALL=False NOPLOTS=True M=$M K=30 N_JOBS=-1 N=$N T=10000 make moremultiplayers; echo "Done for N=$N and M=$M"; read; done; echo "Done for N=$N"; read; done
%
\begin{table}[ht]
  \begin{small}  % XXX
      \centering
      \begin{tabular}{cc|ccc}
          \textbf{Algorithm} & \textbf{Index policy} & $M=3$ players & $M=6$ & $M=9$ \\
          \hline
          \multirow{2}{*}{Centralized multiple-play}
          & \UCB{} & $321 \pm 30$ & $233 \pm 25$ & $0$ \\
          & \klUCB{} & $94 \pm 17$ & $68 \pm 18$ & $0$ \\
          \hline
          \multirow{2}{*}{\Selfish}
          & \UCB{} & $1263 \pm 157$ & $3694 \pm 387$ & $12420 \pm 404$ \\
          & \klUCB{} & $243 \pm 31$ & $743 \pm 113$ & $3005 \pm 492$ \\
          \hline
          \multirow{2}{*}{\RhoRand}
          & \UCB{} & $1455 \pm 208$ & $4775 \pm 463$ & $11794 \pm 1083$ \\
          & \klUCB{} & $394 \pm 96$ & $2385 \pm 412$ & $7057 \pm 1053$ \\
          \hline
          \multirow{2}{*}{\RandTopM}
          & \UCB{} & $1020 \pm 92$ & $2899 \pm 418$ & $470 \pm 346$ \\
          & \klUCB{} & $258 \pm 43$ & $902 \pm 234$ & $551 \pm 520$ \\
          \hline
          \multirow{2}{*}{\MCTopM}
          & \UCB{} & $980 \pm 76$ & $1466 \pm 132$ & $43 \pm 13$ \\
          & \klUCB{} & $\mathbf{248 \pm 40}$ & $\mathbf{410 \pm 54}$ & $\mathbf{42 \pm 10}$ \\
          % \hline
      \end{tabular}
      \caption{For all orthogonalization policies and different numbers of players, using \klUCB{} is much more efficient than using \UCB{}, for multi-players bandit (here in a simple problem with $K=9$ arms).}
      \label{table:5:comparisonUCB_klUCB}
  \end{small}  % XXX
  \end{table}

>>>>

\begin{itemize}
  \item[\modelun]
    \textbf{Simultaneous sensing and collision}: player $j$ observes  $Y_{A^j(t),t}$ \emph{and} $C^j(t)$.
    We note that this first model was never previously studied, but we do not focus on it because of its unrealistic aspect, and its simplicity (it is easier than the following model, for which we obtain good theoretical results).
  \item[\modeldeux]
    \textbf{Sensing, then collision}: player $j$ observes $Y_{A^j(t),t}$, \emph{then} observes the reward, and thus also $C^j(t)$ only if $Y_{A^j(t),t} = 1$.
    This common setup, studied for example by \cite{Anandkumar11,Rosenski16}, is relevant to model the OSA problem: the device first checks for the presence of primary users in the chosen channel,
    if this channel is free ($Y_{A^j(t),t}=1$), the transmission is successful ($r^j(t)=1$) if no collision occurs with other smart devices ($\overline{C^j(t)}$).
  \item[\modeltrois]
    \textbf{No sensing}: player $j$ only observes the reward $r^j(t)$.
    For IoT networks, this reward can be interpreted as an acknowledgement from a Base Station,
    received when a communication was successful.
    A lack of acknowledgment may be due to a collision
    with a device from the background traffic $(Y_{A^j(t),t}=0)$,
    or to a collision with one of the other players ($C^j(t)$).
    However, the sensing and collision information is censored.
    Recently, our work presented in Chapter~\ref{chapter:4} \cite{Bonnefoi17} presented the first (bandit-based) algorithmic solutions under this (harder) feedback model, in a slightly different setup, more suited to large scale IoT applications. As explained above, this third model was found to be too general to analyze mathematically, and we did not present theoretical results of convergence of the algorithms considered in Chapter~\ref{chapter:4}.
    %\footnote{\cite{Bonnefoi17} considers a different setting, more suited for large-scale IoT networks, and a future work will be to extend our work to this setting with more than $K$ players but small probability of activations.}.
\end{itemize}

@@

The regret decomposition in Lemma~\ref{lem:5:DecompositionRegret} is valid for both centralized and decentralized algorithms.
For centralized algorithms, due to the absence of collisions, \ref{eq:5:term3} is obviously zero, and \ref{eq:5:term2} is non-negative, as $N_k(T) \leq T$. For decentralized algorithms, \ref{eq:5:term3} may be significantly large, and term \ref{eq:5:term2} may be negative, as many collisions on arm $k$ may lead to $N_k(T) > T$ (which is counter intuitive with such notations).
However, a careful manipulation of this decomposition shows that the regret is always lower bounded by term \ref{eq:5:term1}.
% The Figure~\ref{fig:5:MP__M9_K9_T10000_N1000__9_algos__main_RegretCentralized____env6} in Appendix illustrate two cases of $M<K$ and $M=K$ and the different impact of the three terms \ref{eq:5:term1}, \ref{eq:5:term2} and \ref{eq:5:term3} on the regret.
Figure~6 in Appendix~F of \cite{Besson2018ALT} illustrates two cases of $M<K$ and $M=K$, and the different impact of the three terms \ref{eq:5:term1}, \ref{eq:5:term2} and \ref{eq:5:term3} on the regret.

@@

Note that $\widetilde{\mu_k}^j(t)$ is no longer meant to be an unbiased estimate of $\mu_k$ as it also takes into account the collision information, that is present in the reward. Based on this empirical reward, one can similarly defined modified indices as (where $f(t)=\mathcal{O}(\log(t))$ is an exploration function)
%
\begin{equation}\label{eq:5:indexTilde}
  \widetilde{U_k}^j(t) \eqdef \begin{cases}
      \widetilde{\mu_k}^j(t)  + \sqrt{  f(t) / (2N_k^j(t))}
      &\text{for } \UCB, \\
      \sup\left\{ q \in [0, 1]: N_k^j(t)\times \kl(\widetilde{\mu_k}^j(t), q) \leq f(t) \right\}
      &\text{for } \klUCB.
  \end{cases}
\end{equation}

@@

% \textbf{Index policies.}
% %
The approaches we now describe for multi-players bandits can be used in combination with any index policy (see Algorithm~\ref{algo:2:indexPolicy}), but we restrict our presentation to \UCB{} algorithms, for which strong theoretical guarantees can be obtained. In particular, we focus on two types of indices:
$\UCB_1$ \citep{Auer02}
and \klUCB{} \citep{KLUCBJournal}.
For player $j\in[M]$
%
denote $S_k^j(t) \eqdef \sum_{s=1}^t Y_{k,s} \indic(A^j(t) = k)$ the current sum of rewards obtained for arm $k$,
then $\widehat{\mu}_k^j(t) \eqdef S_k^j(t)/N_k^j(t)$ (if $N_k^j(t)\neq 0$) is the empirical mean of arm $k$, and thus one can define the index
\begin{equation}\label{eq:5:indexFor_UCB_klUCB}
  U_k^j(t) \eqdef \begin{cases}
      \widehat{\mu}_k^j(t)  + \sqrt{  f(t) / (2N_k^j(t))}
      &\text{for } \UCB, \\
      \sup\left\{ q \in [0, 1]: N_k^j(t)\times \kl(\widehat{\mu}_k^j(t), q) \leq f(t) \right\}
      &\text{for } \klUCB,
  \end{cases}
\end{equation}
where $f(t)$ is some \emph{exploration function}, usually taken to be $\log(t)$ in practice, and slightly larger in theory (\eg, $f(t)=\log(t)+3\log(\log(t))$), which ensures that  $\Pr(U_k^j(t) \geq \mu_k) \gtrsim 1 - 1/t$ (see \cite{KLUCBJournal}).
A classical (single-player) UCB algorithm aims at the arm with the largest index, as presented above in Section~\ref{sub:2:IndexPolicies}.
However, if each of the $M$ players selects the arm with the largest UCB, all the players will end up colliding most of the time on the best arm.
To circumvent this problem, several coordination mechanisms have emerged, that rely on \emph{ordering} the indices and targeting \emph{one of} the $M$-best indices.

@@

\paragraph{The \RandTopM{} algorithm.}
%
We precisely state the first proposal below in Algorithm~\ref{algo:5:RandTopM}.
%
\RandTopM{} is essentially a refinement over \RhoRand, to not use the indirection of ranks, and a simpler version of \MCTopM.
The difference with \MCTopM{} is that the latter introduces a concept of a ``Chair'', by considering a binary ``being fixed'' state $s^j(t)$, as presented in Algorithm~\ref{algo:5:MCTopM} below.
%
In \RandTopM{}, player $j$ is always considered ``not fixed'',
and a \emph{collision always forces a uniform sampling of the next arm} from $\TopM(t)$.

@@

% for N in 4 100; do for M in 3 6 9; do DEBUG=True SAVEALL=False NOPLOTS=True M=$M K=30 N_JOBS=-1 N=$N T=10000 make moremultiplayers; echo "Done for N=$N and M=$M"; read; done; echo "Done for N=$N"; read; done
%
\begin{table}[ht]
  \begin{small}  % XXX
      \centering
      \begin{tabular}{cc|ccc}
          \textbf{Algorithm} & \textbf{Index policy} & $M=3$ players & $M=6$ & $M=9$ \\
          \hline
          \multirow{2}{*}{Centralized multiple-play}
          & \UCB{} & $321 \pm 30$ & $233 \pm 25$ & $0$ \\
          & \klUCB{} & $94 \pm 17$ & $68 \pm 18$ & $0$ \\
          \hline
          \multirow{2}{*}{\Selfish}
          & \UCB{} & $1263 \pm 157$ & $3694 \pm 387$ & $12420 \pm 404$ \\
          & \klUCB{} & $243 \pm 31$ & $743 \pm 113$ & $3005 \pm 492$ \\
          \hline
          \multirow{2}{*}{\RhoRand}
          & \UCB{} & $1455 \pm 208$ & $4775 \pm 463$ & $11794 \pm 1083$ \\
          & \klUCB{} & $394 \pm 96$ & $2385 \pm 412$ & $7057 \pm 1053$ \\
          \hline
          \multirow{2}{*}{\RandTopM}
          & \UCB{} & $1020 \pm 92$ & $2899 \pm 418$ & $470 \pm 346$ \\
          & \klUCB{} & $258 \pm 43$ & $902 \pm 234$ & $551 \pm 520$ \\
          \hline
          \multirow{2}{*}{\MCTopM}
          & \UCB{} & $980 \pm 76$ & $1466 \pm 132$ & $43 \pm 13$ \\
          & \klUCB{} & $\mathbf{248 \pm 40}$ & $\mathbf{410 \pm 54}$ & $\mathbf{42 \pm 10}$ \\
          % \hline
      \end{tabular}
      \caption{Regret for all orthogonalization policies and different numbers of players. Using \klUCB{} is much more efficient than using \UCB{}, for multi-players bandit (here in a simple problem with $K=9$ arms).}
      \label{table:5:comparisonUCB_klUCB}
  \end{small}  % XXX
  \end{table}
